Properties and Bayes fitting of restricted Boltzmann machines

### 2017-07-11 writing/paper-rbm/chapters/rbm.Rmd
edits from Dr. Vardeman

While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning. For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for cell data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However,  the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitted methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. We examine three Bayesian fitting techniques intended to avoid model impropriety in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are closely tied to the probability properties of the RBM model for explaining data generation and it is hoped that the current exposition contributes to a better understanding in this direction.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast; @salakhutdinov2012efficient; @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on, until the desired multi-layer architecture is created.

## Degeneracy, instability, and uninterpretability
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues in model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these characteristics, consider how to quantify these in a RBM, and point out relationships among them.

@@

In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above. @handcock2003assessing classifies an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) (\#eq:elpr)
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for the sequence of RBM models to be stable. That is, we make the following convention.
```{definition, name = "S-unstable RBM", echo=TRUE}
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
```
In other words, the RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

@@

In other words, if the probability ratio (\@ref(eq:elpr)) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Such instability can be a concern in a model for the similar reasons to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains. Recall the applied example of RBM models as a means to classify images. For data as pixels in an image, the instability result in Proposition \@ref(prp:instab1) manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.

@@

The results of this numerical study are summarized in Figures \@ref(fig:degen-plots), \@ref(fig:instab-plots), and \@ref(fig:uninterp-plots). From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \@ref(background). As a compounding issue, these figures show that, as model grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability. Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models. And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety. \par

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.elpr

@@

```
These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space with large $\|\boldsymbol \theta\|$. These problematic aspects require serious consideration when using RBM models, on top of the additional matter of principled/rigorous fitting of RBM models.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\@ref(eq:pmf)) of any realistic size. Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).

@@

for the outcomes $(\pm 1, \pm 1)$, the appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the appendix shows, only it is possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated. For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

For both BwTNLV and BwTNML methods, effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \@ref(tab:m-eff). These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \@ref(fig:acf), BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.

@@

Ultimately, it is not clear that RBM models are useful as generative models. Furthermore, without the appropriate generative behavior in the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions). In the case of classification, predictive distributions not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.  For these reasons, we are skeptical about RBMs as data generative tools.

>>>>

While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning. For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions. These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.

@@

RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast; @salakhutdinov2012efficient; @srivastava2013modeling for examples]. By stacking multiple layers of RBMs in a deep architecture, proponents of the model are claimed to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp. 450]. The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on, until the desired multi-layer architecture is created.

## Degeneracy, instability, and uninterpretability
The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*. In this section we define these, consider how to quantify them in a RBM, and point out relationships among them.

@@

In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution with a probability mass function of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above. @handcock2003assessing classifies an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.

@@

Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*. Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$. Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability. To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles. When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used). To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) (\#eq:elpr)
\end{align}
where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for a sequence of RBM models to be stable. That is, we make the following convention.
```{definition, name = "S-unstable RBM", echo=TRUE}
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
```
In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$. This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.

@@

In other words, if the probability ratio (\@ref(eq:elpr)) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains. Recall the applied example of RBM models as a means to classify images. For data as pixels in an image, the instability result in Proposition \@ref(prp:instab1) manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety (near-degeneracy, instability, and uninterpretability), for RBM models of varying sizes.

@@

The results of this numerical study are summarized in Figures \@ref(fig:degen-plots), \@ref(fig:instab-plots), and \@ref(fig:uninterp-plots). From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \@ref(background). As a compounding issue, these figures show that, as models grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability. Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models. And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety. \par

@@

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), V))), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.elpr

@@

```
These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space having large $\|\boldsymbol \theta\|$. These problematic aspects require serious consideration when using RBM models, on top of the additional matter of principled/rigorous fitting of RBM models.

@@

Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\@ref(eq:pmf)) of any realistic size. Ad hoc methods are often recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).

@@

for the outcomes $(\pm 1, \pm 1)$, the appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated. For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

For both BwTNLV and BwTNML methods, effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \@ref(tab:m-eff). These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations of the MCMC chain to be run in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \@ref(fig:acf), BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.

@@

Ultimately, it is not clear that RBM models are useful as generative models. Furthermore, without the appropriate generative behavior in the RBM model, quantifying uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions). In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.  For these reasons, we are skeptical about RBMs as data generative tools.
